---
title: "initial_gene_exp_analysis"
output: html_document
date: "2025-05-29"
---

## Extracting Expression Data from OMIM_data

Loading in the two separate OMIM datasets:

1.  General macrocephaly genes
2.  Other brain enlargement disorders - Sotos and Malan syndrome

```{r}
#Loading in dataset
OMIM_dat <- read.delim("../data/OMIM_Databases_Gene_Map_Queries/cleaned OMIM_data_macro.tsv", sep="\t", na="NA")
View(OMIM_dat)
print(rownames(seurat_subset_cleaned))
extracting_genes <- intersect(rownames(seurat_subset_cleaned), OMIM_dat$Mouse.Gene..from.MGI.) #Don't need to convert anything - we had a whole column for it - 175 genes
View(extracting_genes)
print(extracting_genes)

#Loading in second OMIM dataset 
OMIM_dat_2 <- read.delim("../data/OMIM_Databases_Gene_Map_Queries/cleaned_OMIM_data_MALAN_SOTOS.tsv", sep = "\t", na = "NA")
head(OMIM_dat_2)

malan_sotos_genes <- intersect(rownames(seurat_subset_cleaned), OMIM_dat_2$Mouse.Gene..from.MGI.)
View(malan_sotos_genes)
```

## Initial Downstream Analysis and Data Visualization of Macrocephaly Genes 

### 1. Plot Heatmap of Expression of Macrocephaly Genes

#### AverageExpression vs AggregateExpression

```{r}

print(intersect(extracting_genes, rownames(seurat_subset_cleaned)))
avg_exp <- AverageExpression(
  seurat_subset_cleaned, features = extracting_genes, group.by = "New_cellType", slot = "data"
)

avg_exp_mat <- as.matrix(avg_exp$RNA)
dim(avg_exp_mat) #[1] 175 24


#Apply the z_score normalization function to the genes (rows)
z_avg_exp <- t(apply(avg_exp_mat, 1, function(x) (x-mean(x))/sd(x))) ##Makes sure the data is reformatted back to genes as rows 

#Used to check if z-score has been correctly calculated 
print(mean(z_avg_exp[1,])) #[1] 1.387779e-17- Close to zero 
print(mean(z_avg_exp[,1])) #[1] NaN

#Doing final check that the z-score normalization was ran correctly 
sd_row <- apply(z_avg_exp, 1, sd) #All have an SD of 1 
mean_row <- apply(z_avg_exp,1, mean) #All have a mean value clos
mean_col <- apply(z_avg_exp, 2, mean) #All NaN for Cell Types - SHOULD BE 


#Checking for any non-variance genes or NaN values BEFORE z-score normalization
error_genes <- apply(z_avg_exp, 1, function (x){
  #Check for NAs
  any(is.na(x)) ||
  
  #Check for infinite vals
  any(is.infinite(x)) ||
  
  #Check for zero SD (Non-var)
  sd(x, na.rm = TRUE) == 0
})

z_avg_exp_filtered <- z_avg_exp[!error_genes,] #Filter out genes that are flagged in the function
print(dim(z_avg_exp_filtered))

############ Plotting z_score_normalized graphs 

########## Graph #1 using Pretty Heatmap

install.packages("pheatmap")
library(pheatmap)
library(RColorBrewer)


avg_exp_z_heatmap <- pheatmap(z_avg_exp_filtered, color = colorRampPalette(brewer.pal(5, "RdPu"))(100), clustering_method = "complete", cluster_rows = TRUE, cluster_cols = TRUE, fontsize_row = 8, fontsize_col = 8, angle_col = 45, raster = TRUE, annotation_legend_param = list(title = "Z-score Normalization"),filename = "../outputs/z_score_norm_heatmaps/avg_exp_z_score_macro_pheatmap.png")

print(avg_exp_z_heatmap)


###TRY AND MAKE IT WORK TO INCLUDE A LABEL FOR THE COLOUR BAR 

library(reshape2)
library(ggplot2)

# Assuming z_agg_exp_filtered is a matrix
df <- melt(z_avg_exp_filtered)
names(df) <- c("Row", "Column", "Value")

# Create a ggplot heatmap
p <- ggplot(df, aes(x = Column, y = Row, fill = Value)) +
  geom_tile() +
  scale_fill_gradientn(colors = colorRampPalette(brewer.pal(5, "RdPu"))(100),
                      name = "Z-score\nNormalization") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y = element_text(size = 8)) +
  coord_fixed()

# Save the plot
ggsave("../outputs/z_score_norm_heatmaps/avg_exp_z_score_macro_ggplot.png", p, width = 30, height = 20, units = "in", dpi = 100)

test_color <- colorRampPalette(brewer.pal(5,"RdPu"))(100)
print(test_color)
```

**As of v5, Seurat recommends using AggregateExpression over AverageExpression**

```{r}

intersect(extracting_genes, rownames(seurat_subset))

##Altnernative method using AggregateExpression
agg_exp <- AggregateExpression(
  seurat_subset_cleaned, features = extracting_genes, group.by = "New_cellType", slot = "data"
) #Seurat recommended using AggregateExpression

agg_exp_mat <- as.matrix(agg_exp$RNA)
dim(agg_exp_mat)

#Apply the z_score normalization function to the genes (rows)
z_agg_exp <- t(apply(agg_exp_mat, 1, function(x) (x-mean(x))/sd(x)))

#Used to check if z-score has been correctly calculated 
print(mean(z_agg_exp[1,])) #[1] -2.659909e-17 - Close to zero 
print(mean(z_agg_exp[,1])) #[1] NaN

#Doing final check that the z-score normalization was ran correctly 
sd_row <- apply(z_agg_exp, 1, sd) #All have an SD of 1 
mean_row <- apply(z_agg_exp,1, mean) #All have a mean value clos
mean_col <- apply(z_agg_exp, 2, mean) #All NaN for Cell Types - SHOULD BE 


#Checking for any non-variance genes or NaN values BEFORE z-score normalization
error_genes <- apply(z_agg_exp, 1, function (x){
  #Check for NAs
  any(is.na(x)) ||
  
  #Check for infinite vals
  any(is.infinite(x)) ||
  
  #Check for zero SD (Non-var)
  sd(x, na.rm = TRUE) == 0
})

z_agg_exp_filtered <- z_agg_exp[!error_genes,] #Filter out genes that are flagged in the function
print(dim(z_agg_exp_filtered))


install.packages("pheatmap")
library(pheatmap)
library(RColorBrewer)

agg_exp_z_heatmap <- pheatmap(z_agg_exp_filtered, color = colorRampPalette(brewer.pal(5, "RdPu"))(100), clustering_method = "complete", cluster_rows = TRUE, cluster_cols = TRUE, fontsize_row = 8, fontsize_col = 8, angle_col = 45, raster = TRUE, annotation_legend_param = list(title = "Z-score Normalization"),filename = "../outputs/z_score_norm_heatmaps/agg_exp_z_score_macro_pheatmap.png")

print(agg_exp_z_heatmap)
View()


###TRY AND MAKE IT WORK TO INCLUDE A LABEL FOR THE COLOUR BAR 

library(reshape2)
library(ggplot2)

# Assuming z_agg_exp_filtered is a matrix
df <- melt(z_agg_exp_filtered)
names(df) <- c("Row", "Column", "Value")

# Create a ggplot heatmap
p <- ggplot(df, aes(x = Column, y = Row, fill = Value)) +
  geom_tile() +
  scale_fill_gradientn(colors = colorRampPalette(brewer.pal(5, "RdPu"))(100),
                      name = "Z-score\nNormalization") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y = element_text(size = 8)) +
  coord_fixed()

# Save the plot
ggsave("../outputs/z_score_norm_heatmaps/agg_exp_z_score_macro_ggplot.png", p, width = 30, height = 20, units = "in", dpi = 100)

test_color <- colorRampPalette(brewer.pal(5,"RdPu"))(100)
print(test_color)


```

##### Confirming which Heatmap is correct by using median expression of each gene across all cell types and then performing z-score normalization

```{r}

gene_exp <- GetAssayData(seurat_subset_cleaned, assay = "RNA", slot = "data")#Pulling out the normalized 'data' slot/layer 
gene_exp_df <- as.data.frame(gene_exp)
gene_exp_macro <- subset(gene_exp_df, rownames(gene_exp_df) %in% extracting_genes) #subsetting to include data of only macrocephaly gene

View(gene_exp_macro)

#Extracting cell type information from meta.data 
cell_types <- seurat_subset_cleaned@meta.data$New_cellType
names(cell_types) <- colnames(seurat_subset_cleaned)

unique_cell_types <- unique(cell_types)
genes_by_cell_types <- matrix(0, 
                              nrow = nrow(gene_exp_macro), 
                              ncol = length(unique_cell_types),
                              dimnames = list(rownames(gene_exp_macro), unique_cell_types))

#Calculating median expression for each gene in EACH cell type 
for (cell_type in unique_cell_types) {
  cells_of_type <- names(cell_types)[cell_types == cell_type]
  
  #Calculate median expression for each gene in this cell type 
  genes_by_cell_types[, cell_type] <- apply(gene_exp_macro[, cells_of_type, drop = FALSE], 1, median, na.rm = TRUE)
}

#Conver to dataframe 
med_exp_gene_in_cell_type_df <- as.data.frame(genes_by_cell_types)

View(head(med_exp_gene_in_cell_type_df))


######## Try Plotting median expression as it is without the z-score normalization
library(ggplot2)
library(reshape2)

# Convert dataframe to long format
long_df <- melt(as.matrix(med_exp_gene_in_cell_type_df))

# Create plot
median_heatmap <- ggplot(long_df, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile() +
  scale_fill_gradientn(colors = colorRampPalette(brewer.pal(5, "RdPu"))(100),
                      name = "Z-score\nNormalization") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y = element_text(size = 8)) +
  coord_fixed()

ggsave("median_pre_normalization.png", median_heatmap, width = 30, height = 30)

#Performing z-score normalization

########### Approach 1 
z_score_med_exp <- t(apply(med_exp_gene_in_cell_type_df, 1, function(x) (x-mean(x))/sd(x)))

########### Approach 2 
library(matrixStats)
gene_means <- rowMeans(med_exp_gene_in_cell_type_df, na.rm = TRUE)
gene_sds <- rowSds(as.matrix(med_exp_gene_in_cell_type_df), na.rm = TRUE)

z_score_df <- sweep(med_exp_gene_in_cell_type_df, 1, gene_means, "-")
z_score_df <- sweep(z_score_df, 1, gene_sds, "/")


########### Approach 3
avg_median_per_gene <- rowMeans(med_exp_gene_in_cell_type_df, na.rm = TRUE)
sd_median_per_gene <- apply(med_exp_gene_in_cell_type_df, 1, sd, na.rm = TRUE)

z_score_df_2 <- med_exp_gene_in_cell_type_df

for (i in 1:nrow(z_score_df_2)) {
  if (is.na(sd_median_per_gene[i]) || sd_median_per_gene[i] == 0) {
    z_score_df_2[i, ] <- 0 
  } else {
    z_score_df_2[i,] <- (med_exp_gene_in_cell_type_df[i,] - avg_median_per_gene[i]) / sd_median_per_gene[i]
  }
}

########## Appraoch 4 - Using MAD 
median_mad_normalize <- function(df) {
  normalize_df <- df #Create empty dataframe 
  
  for (gene in rownames(df)) {
    values <- as.numeric(df[gene, ])
    
    gene_median <- median(values, na.rm = TRUE)
    gene_MAD <- mad(values, na.rm = TRUE, constant = 1)
    
    if(gene_MAD == 0) {
      normalize_df[gene, ] <- 0
    } else {
      normalize_df[gene, ] <- (values - gene_median) / gene_MAD
    }
  }
  
  return(normalize_df)
} 

z_score_MAD_df <- median_mad_normalize(med_exp_gene_in_cell_type_df)

#Final Checks and removing non-zero variance values
#Used to check if z-score has been correctly calculated 
print(mean(z_score_med_exp[1,])) #[1] Is returning an NA - Need to check why 
print(mean(z_score_med_exp[,1])) #[1] NaN

#Doing final check that the z-score normalization was ran correctly 
sd_row <- apply(z_score_med_exp, 1, sd) #Either returning an NA or 1 
mean_row <- apply(z_score_MAD_df,1, mean) #Either returning an NA or a value close to zero 
mean_col <- apply(z_score_df, 2, mean) #All NaN for Cell Types - SHOULD BE 


#Checking for any non-variance genes or NaN values AFTER z-score normalization
error_genes <- apply(z_score_med_exp, 1, function (x){
  #Check for NAs
  any(is.na(x)) ||
  
  #Check for infinite vals
  any(is.infinite(x)) ||
  
  #Check for zero SD (Non-var)
  sd(x, na.rm = TRUE) == 0
})

z_score_med_filtered <- z_score_med_exp[!error_genes, ]

z_score_df_filtered <- z_score_df[!error_genes,] #Filter out genes that are flagged in the function
print(dim(z_score_df_filtered))

z_score_df_2_filtered <- z_score_df_2[!error_genes, ]


z_score_med_exp_heatmap <- pheatmap(z_score_med_filtered, color = colorRampPalette(brewer.pal(5, "RdPu"))(100), clustering_method = "complete", cluster_rows = TRUE, cluster_cols = TRUE, fontsize_row = 8, fontsize_col = 8, angle_col = 45, raster = TRUE, annotation_legend_param = list(title = "Z-score Normalization"),filename = "median_exp_z_score_macro_pheatmap_manual.png")
```

Median was not an effective method - too many zero expression values

### 2. Plot Heatmap of Other Macro Genes

#### AverageExpression 

```{r}
print(intersect(malan_sotos_genes, rownames(seurat_subset_cleaned)))
avg_exp <- AverageExpression(
  seurat_subset_cleaned, features = malan_sotos_genes, group.by = "New_cellType", slot = "data"
)

avg_exp_mat <- as.matrix(avg_exp$RNA)
dim(avg_exp_mat) #[1] 13 24


#Apply the z_score normalization function to the genes (rows)
z_avg_exp <- t(apply(avg_exp_mat, 1, function(x) (x-mean(x))/sd(x))) ##Makes sure the data is reformatted back to genes as rows 

#Used to check if z-score has been correctly calculated 
print(mean(z_avg_exp[1,])) #[1] -6.667e-18- Close to zero 
print(mean(z_avg_exp[,1])) 

#Doing final check that the z-score normalization was ran correctly 
sd_row <- apply(z_avg_exp, 1, sd) #All have an SD of 1 
mean_row <- apply(z_avg_exp,1, mean) 
mean_col <- apply(z_avg_exp, 2, mean)  


#Checking for any non-variance genes or NaN values BEFORE z-score normalization
error_genes <- apply(z_avg_exp, 1, function (x){
  #Check for NAs
  any(is.na(x)) ||
  
  #Check for infinite vals
  any(is.infinite(x)) ||
  
  #Check for zero SD (Non-var)
  sd(x, na.rm = TRUE) == 0
})

z_avg_exp_filtered <- z_avg_exp[!error_genes,] #Filter out genes that are flagged in the function
print(dim(z_avg_exp_filtered)) #[1] 13 24 - No errors 

############ Plotting z_score_normalized graphs 

########## Graph #1 using Pretty Heatmap

install.packages("pheatmap")
library(pheatmap)
library(RColorBrewer)


avg_exp_z_heatmap <- pheatmap(z_avg_exp_filtered, color = colorRampPalette(brewer.pal(5, "RdPu"))(100), clustering_method = "complete", cluster_rows = TRUE, cluster_cols = TRUE, fontsize_row = 8, fontsize_col = 8, angle_col = 45, raster = TRUE, annotation_legend_param = list(title = "Z-score Normalization"),filename = "../outputs/z_score_norm_heatmaps/avg_exp_z_score_malan_and_sotos_pheatmap.png")

print(avg_exp_z_heatmap)


###TRY AND MAKE IT WORK TO INCLUDE A LABEL FOR THE COLOUR BAR 

library(reshape2)
library(ggplot2)

# Assuming z_agg_exp_filtered is a matrix
df <- melt(z_avg_exp_filtered)
names(df) <- c("Row", "Column", "Value")

# Create a ggplot heatmap
p <- ggplot(df, aes(x = Column, y = Row, fill = Value)) +
  geom_tile() +
  scale_fill_gradientn(colors = colorRampPalette(brewer.pal(5, "RdPu"))(100),
                      name = "Z-score\nNormalization") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
        axis.text.y = element_text(size = 15)) +
  coord_fixed()

# Save the plot
ggsave("../outputs/z_score_norm_heatmaps/avg_exp_z_score_malan_and_sotos_ggplot.png", p, width = 30, height = 20, units = "in", dpi = 100)

test_color <- colorRampPalette(brewer.pal(5,"RdPu"))(100)
print(test_color)
```
