---
title: "OMIM_Dataset_Final"
output: html_document
date: "2025-03-09"
---

## Loading two OMIM tsv files

-   **One file is a dataset that was generated using the search term 'macrocephaly' and the second one is one that was generated using the search term 'macrocephaly, NE-RG transition'**
-   **For all downstream cleaning, I only applied it to the 'macrocephaly' dataset**

```{r}
getwd()

##Loading in OMIM table queried by 'macrocephaly'
OMIM_dat <- read.delim("OMIM_Databases_Gene_Map_Queries/OMIM-Gene-Map-Retrieval.tsv", header = TRUE, sep = "\t", check.names = TRUE, skip =3)
View(OMIM_dat)
dim(OMIM_dat) #[1] 376 14 - 376 entries 
```

```{r}
##Loading in OMIM table queried by 'macrocephaly, NE-RG transition'
OMIM_dat_NE_RG <- read.delim("OMIM_Databases_Gene_Map_Queries/OMIM-Gene-Map-Retrieval-NE-RG.tsv", header =TRUE, sep="\t", check.names = TRUE, skip =3)
View(OMIM_dat_NE_RG)
dim(OMIM_dat_NE_RG) #[1] 307 14 - 307 entries 
```

```{r}
#Installing cleaning packages 
install.packages("dplyr")
library(dplyr)
install.packages("tidyr")
library(tidyr)
```

## Cleaning Datasets

#### 1. Removing the pheno map key description at the bottom

```{r}

OMIM_dat <- OMIM_dat[1:349, ] #Last entry was row 349

View(OMIM_dat)
```

#### 2. Moving misaligned entries

**A couple of the Mouse Gene names in the last column are in the Pheno Map Key column**

```{r}
print(OMIM_dat[,13:14])

```

``` r
print(OMIM_dat[,13:14])


Pheno.map.key
<chr>
Mouse.Gene..from.MGI.
<chr>
3   Gatad2b         
3   Ash1l           
3   Rit1            
3   Pigm            
3   Tmco1           
3   Cacna1e         
3   Kdm5b           
Srgap2              
3   Syt14
```

```{r}
#First applying it on the OMIM_dat_sep 
typeof(OMIM_dat$Pheno.map.key) #ensuring the Pheno.map.key is a "character variable"

OMIM_dat$Pheno.map.key <- as.character(OMIM_dat$Pheno.map.key)

for (i in 1:nrow(OMIM_dat)) {
  # Check if the value in 'Pheno.map.key' is a character (likely a gene name) and not a number
  if (!is.na(OMIM_dat$Pheno.map.key[i]) && !grepl("^\\d+$", OMIM_dat$Pheno.map.key[i])) {
    # Shift the value (gene name) to 'Mouse.Gene..from.MGI.' column
    OMIM_dat$Mouse.Gene..from.MGI.[i] <- OMIM_dat$Pheno.map.key[i]
    # Set the corresponding value in 'Pheno.map.key' to NA
    OMIM_dat$Pheno.map.key[i] <- NA
  }
}

View(OMIM_dat)
print(subset(OMIM_dat, OMIM_dat$Mouse.Gene..from.MGI. == "Srgap2")) #Srgap2 has now been shifted into the correct column and the corresponding pheno.map.key column in filled in with NA
```

**(How to clean the second dataset, 'OMIM_dat_NE_RG')**

```{r}
print(OMIM_dat_NE_RG[,13:14])

```

``` r

Pheno.map.key
<chr>
Mouse.Gene..from.MGI.
<chr>
2               
2               
2               
3   Isg15           
3   Isg15           
3   Isg15           
3   Isg15           
Mir200b             
Mir200b             
Mir200a
```

```{r}
#Applying it to the second dataset 'OMIM_NE_RG_sep'
typeof(OMIM_dat_NE_RG$Pheno.map.key) #ensuring the Pheno.map.key is a "character variable"

OMIM_dat_NE_RG$Pheno.map.key <- as.character(OMIM_dat_NE_RG$Pheno.map.key)

for (i in 1:nrow(OMIM_dat_NE_RG)) {
  # Check if the value in 'Pheno.map.key' is a character (likely a gene name) and not a number
  if (!is.na(OMIM_dat_NE_RG$Pheno.map.key[i]) && !grepl("^\\d+$", OMIM_dat_NE_RG$Pheno.map.key[i])) {
    # Shift the value (gene name) to 'Mouse.Gene..from.MGI.' column
    OMIM_dat_NE_RG$Mouse.Gene..from.MGI.[i] <- OMIM_dat_NE_RG$Pheno.map.key[i]
    # Set the corresponding value in 'Pheno.map.key' to NA
    OMIM_dat_NE_RG$Pheno.map.key[i] <- NA
  }
}

View(OMIM_dat_NE_RG)
print(subset(OMIM_dat_NE_RG, OMIM_dat_NE_RG$Mouse.Gene..from.MGI. == "Mir200b")) #Mir200b has now been shifted into the correct column and the corresponding pheno.map.key column in filled in with NA


```

#### Unlike the previous Rmd, I am not going to split the Gene.Locus entries - the multiple entries are the other names for that gene (genes have multiple names) 

## Saving datasets as csv files

```{r}
install.packages("readr")
library(readr)
write_delim(OMIM_dat, file = "cleaned OMIM_data_macro.tsv", delim = "\t", na="NA", col_names = TRUE)
write_delim(OMIM_dat_NE_RG, file = "cleaned OMIM_data_macro_NE_RG.tsv", delim = "\t", na="NA", col_names = TRUE)

```
